---
title: "VTPEH 6270 – Check Point 04: Data Simulation"
author: "Asma Darwish"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
---

```{r install-packages, eval=FALSE, include=FALSE}
# One-time setup — never runs automatically during knit
install.packages("kableExtra", repos = "https://cran.r-project.org")
tinytex::reinstall_tinytex(repository = "illinois")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 5.5, fig.height = 3.5)
library(tidyverse)
library(knitr)
library(dplyr)
library(kableExtra)
set.seed(42)
```

---

# Introduction

This report is part of a series of checkpoints for VTPEH 6270, building on prior
data exploration of the U.S. COVID-19 Treatment Locator dataset.
The dataset documents 64,236 provider locations across the United States and
captures which oral antiviral treatments --- Paxlovid (nirmatrelvir/ritonavir),
Lagevrio (molnupiravir), and Veklury (remdesivir) --- are available at each site,
along with service characteristics such as telehealth availability, pharmacist
prescribing authority, and patient assistance program (PAP) enrollment.

In Checkpoint 3, data exploration revealed striking disparities between telehealth
and non-telehealth sites: telehealth sites (n = 390, 0.6% of all sites) showed
substantially higher rates of pharmacist prescribing authority (90.3% vs. 0.0%),
Paxlovid availability (88.2% vs. 18.1%), and Lagevrio availability (100% vs.
13.4%) compared to non-telehealth sites (n = 63,847). These patterns motivated a
refined research question focused specifically on Paxlovid availability, the most
widely used COVID-19 oral antiviral.

This checkpoint focuses on **data simulation** in the context of power analyses,
using the observed data from Checkpoint 3 to parameterize plausible biological
and sampling scenarios for the association between telehealth status and Paxlovid
availability.

---

# Objective and Approach

## Specific Objective

> **Research question:** Are sites that offer telehealth services more likely to
> stock Paxlovid (nirmatrelvir/ritonavir) than sites that do not offer telehealth,
> among U.S. COVID-19 treatment providers?

## Data Description

```{r load-data}
dat <- read.csv("COVID-19_Treatments_20260127.csv", stringsAsFactors = FALSE)
```

```{r variable-table, echo=FALSE}
var_table <- data.frame(
  Variable    = c("Is.Telehealth.Site", "Has.Paxlovid"),
  Type        = c("Binary / Categorical", "Binary / Categorical"),
  Class_in_R  = c("character to logical", "character to logical"),
  Description = c(
    "Whether the provider offers telehealth services (TRUE/FALSE)",
    "Whether the provider currently stocks Paxlovid (TRUE/FALSE)"
  )
)

kable(var_table,
      col.names = c("Variable Name", "Variable Type", "R Class", "Description"),
      caption   = "Variables of interest for the research question.",
      booktabs  = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 9) %>%
  column_spec(4, width = "6cm")
```

```{r recode}
dat <- dat %>%
  mutate(
    telehealth = Is.Telehealth.Site == "TRUE",
    paxlovid   = Has.Paxlovid       == "TRUE"
  )
```

## Data Visualization

```{r viz-barplot, fig.cap="Proportion of providers stocking Paxlovid by telehealth status.", fig.width=4, fig.height=3.2}
dat %>%
  group_by(telehealth) %>%
  summarise(prop_paxlovid = mean(paxlovid), n = n(), .groups = "drop") %>%
  mutate(telehealth_label = ifelse(telehealth, "Telehealth", "Non-Telehealth")) %>%
  ggplot(aes(x = telehealth_label, y = prop_paxlovid, fill = telehealth_label)) +
  geom_col(width = 0.35, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(prop_paxlovid * 100, 1), "%\n(n = ", n, ")")),
            vjust = -0.4, size = 3) +
  scale_fill_manual(values = c("Non-Telehealth" = "#6baed6",
                               "Telehealth"     = "#2171b5")) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.1)) +
  labs(x = NULL, y = "Proportion Stocking Paxlovid",
       title = "Paxlovid Availability by Telehealth Status") +
  theme_bw(base_size = 10) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 10),
        panel.grid.major.x = element_blank())
```

## Description of Plausible Relationships

Visual inspection confirms the pattern identified in Checkpoint 3: telehealth
providers stock Paxlovid at a substantially **higher rate** than non-telehealth
providers (approximately 88% vs. 18%). This is consistent with a positive
association between telehealth capability and antiviral availability, likely
reflecting that better-resourced sites with broader service capacity tend to
offer both features. The relationship is modelled as a **difference in
proportions** between two binary groups (telehealth status and Paxlovid
availability):

$$P(\text{Paxlovid} = 1 \mid \text{Telehealth} = 1) = p_1$$
$$P(\text{Paxlovid} = 1 \mid \text{Telehealth} = 0) = p_0$$
$$\text{Effect size (odds ratio)} = OR = \frac{p_1/(1-p_1)}{p_0/(1-p_0)}$$

## Parameters of Interest

```{r obs-params}
# obs-params runs FIRST so p0_obs, p1_obs, OR_obs exist before param-plot uses them
p0_obs <- mean(dat$paxlovid[!dat$telehealth])
p1_obs <- mean(dat$paxlovid[ dat$telehealth])
OR_obs <- (p1_obs / (1 - p1_obs)) / (p0_obs / (1 - p0_obs))

cat(sprintf("p0 (non-telehealth):         %.3f\n", p0_obs))
cat(sprintf("p1 (telehealth):             %.3f\n", p1_obs))
cat(sprintf("Odds Ratio:                  %.2f\n",  OR_obs))

prop_tele_obs <- mean(dat$telehealth)
cat(sprintf("Proportion telehealth sites: %.4f\n", prop_tele_obs))
```

```{r param-plot, echo=FALSE, fig.width=5.5, fig.height=3, fig.cap="Observed model parameters estimated from the HHS COVID-19 Treatment Locator dataset (n = 64,237 sites, January 2026). OR = odds ratio; divided by 100 for display on a shared axis."}

param_plot_data <- data.frame(
  label    = c("p0  (Non-Telehealth)", "p1  (Telehealth)", "OR  (scaled x0.01)"),
  value    = c(p0_obs, p1_obs, OR_obs / 100),
  display  = c(sprintf("%.3f", p0_obs),
               sprintf("%.3f", p1_obs),
               sprintf("OR = %.1f", OR_obs)),
  category = c("Probability", "Probability", "Effect Size"),
  stringsAsFactors = FALSE
)

param_plot_data$label <- factor(param_plot_data$label,
                                 levels = c("OR  (scaled x0.01)",
                                            "p1  (Telehealth)",
                                            "p0  (Non-Telehealth)"))

ggplot(param_plot_data,
       aes(x = value, y = label, color = category, shape = category)) +
  geom_vline(xintercept = 0, color = "grey80", linewidth = 0.4) +
  geom_segment(aes(x = 0, xend = value, yend = label),
               linewidth = 0.7, alpha = 0.55) +
  geom_point(size = 4.5) +
  geom_text(aes(label = display),
            hjust = -0.2, size = 3, color = "black") +
  scale_color_manual(values = c("Probability" = "#2171b5",
                                "Effect Size"  = "#cb181d"),
                     name   = NULL) +
  scale_shape_manual(values = c("Probability" = 16,
                                "Effect Size"  = 17),
                     name   = NULL) +
  scale_x_continuous(limits = c(0, 1.35),
                     breaks = c(0, 0.25, 0.50, 0.75, 1.0),
                     labels = c("0", "0.25", "0.50", "0.75", "1.0")) +
  labs(
    x     = "Estimated Value  (OR divided by 100 for display)",
    y     = NULL,
    title = "Observed Parameter Estimates"
  ) +
  theme_classic(base_size = 10) +
  theme(
    plot.title         = element_text(face = "bold", size = 10, hjust = 0),
    axis.text.y        = element_text(size = 9,  color = "black"),
    axis.text.x        = element_text(size = 8,  color = "grey30"),
    axis.title.x       = element_text(size = 8,  color = "grey30"),
    axis.line          = element_line(color = "grey60", linewidth = 0.4),
    axis.ticks         = element_line(color = "grey60"),
    legend.position    = c(0.80, 0.25),
    legend.text        = element_text(size = 8),
    legend.key.size    = unit(0.4, "cm"),
    legend.background  = element_rect(fill  = "white",
                                      color = "grey85",
                                      linewidth = 0.3),
    panel.grid.major.x = element_line(color = "grey92", linewidth = 0.3),
    plot.margin        = margin(8, 40, 8, 8)
  )
```

---

# Simulation

## Simulation Basis

The simulation generates binary outcomes for two groups (telehealth vs.
non-telehealth) under a logistic model. Group membership is drawn randomly, and
Paxlovid availability is drawn from Bernoulli distributions with probabilities
$p_0$ (non-telehealth) and $p_1$ (telehealth), derived from the observed data
in Checkpoint 3.

```{r sim-helpers}
# Helper: convert OR + p0 to p1
or_to_p1 <- function(p0, OR) {
  odds0 <- p0 / (1 - p0)
  odds1 <- odds0 * OR
  odds1 / (1 + odds1)
}

# Summary statistics: proportions stocking Paxlovid by telehealth group
sim_summary <- function(df) {
  df %>%
    group_by(telehealth) %>%
    summarise(n             = n(),
              n_paxlovid    = sum(paxlovid),
              prop_paxlovid = mean(paxlovid),
              .groups = "drop")
}
```

## Simulation Function

```{r sim-function}
#' Simulate COVID-19 Paxlovid availability data and return summary statistics.
#' @param n          Total sample size
#' @param OR         Odds ratio (effect size); OR = 1 means no association
#' @param noise      SD of log-odds noise added to baseline probability
#' @param p0         Baseline probability of Paxlovid at non-telehealth sites
#' @param prop_tele  Proportion of sites that are telehealth
#' @return           A list with `data` (raw data frame) and `summary` (summary stats)
simulate_paxlovid <- function(n, OR, noise, p0 = 0.18, prop_tele = prop_tele_obs) {
  p0_noisy <- plogis(qlogis(p0) + rnorm(1, 0, noise))
  p0_noisy <- pmax(0.01, pmin(0.99, p0_noisy))
  p1       <- or_to_p1(p0_noisy, OR)

  telehealth <- rbinom(n, 1, prop_tele)
  paxlovid   <- ifelse(telehealth == 1,
                       rbinom(n, 1, p1),
                       rbinom(n, 1, p0_noisy))
  df <- data.frame(telehealth = as.logical(telehealth),
                   paxlovid   = as.logical(paxlovid))
  list(data    = df,
       summary = sim_summary(df))
}
```

## Simulation Output

```{r sim-output}
example_sim <- simulate_paxlovid(n = 1000, OR = 30, noise = 0)
kable(example_sim$summary, digits = 3,
      caption  = "Example simulation output (n = 1000, OR = 30).",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```

## Simulation Automation

```{r sim-automation, cache=TRUE}
effect_sizes <- c(1, 2, 3, 5, 8, 12, 20, 30, 50, 80)
sample_sizes <- c(100, 300, 500, 1000, 2000, 5000,
                  10000, 20000, 40000, 64000)
noise_levels <- c(0, 0.5, 1.5)

sim_grid <- expand.grid(n     = sample_sizes,
                        OR    = effect_sizes,
                        noise = noise_levels)

set.seed(42)
sim_results <- vector("list", nrow(sim_grid))
for (i in seq_len(nrow(sim_grid))) {
  sim_results[[i]] <- simulate_paxlovid(
    n     = sim_grid$n[i],
    OR    = sim_grid$OR[i],
    noise = sim_grid$noise[i]
  )
}

cat(sprintf("Total simulations run: %d\n", length(sim_results)))
```

---

# Visualization

```{r viz-simulation, fig.width=7, fig.height=3.8, fig.cap="Simulated proportion of telehealth sites stocking Paxlovid across effect sizes and sample sizes, faceted by noise level. Higher OR = larger true effect; larger n = more stable estimates."}

extract_tele_prop <- function(sim_list) {
  s   <- sim_list$summary
  row <- s[s$telehealth == TRUE, ]
  if (nrow(row) == 0) return(NA_real_)
  row$prop_paxlovid
}

sim_grid$prop_tele <- sapply(sim_results, extract_tele_prop)
n_missing <- sum(is.na(sim_grid$prop_tele))
if (n_missing > 0)
  message(sprintf("Note: %d runs had no telehealth sites and are omitted.", n_missing))

sim_grid$noise_label <- factor(paste0("Noise SD = ", sim_grid$noise),
                                levels = paste0("Noise SD = ", noise_levels))
sim_grid$n_label <- factor(sim_grid$n)

ggplot(sim_grid, aes(x = factor(OR), y = prop_tele,
                     color = n_label, group = n_label)) +
  geom_line(alpha = 0.5, linewidth = 0.4) +
  geom_point(size = 0.9, alpha = 0.7, na.rm = TRUE) +
  facet_wrap(~ noise_label, ncol = 3) +
  scale_color_viridis_d(name = "n", option = "C",
                        guide = guide_legend(nrow = 1,
                                             override.aes = list(size = 2,
                                                                 linewidth = 0.8))) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1),
                     breaks = c(0, 0.5, 1)) +
  labs(
    x     = "Odds Ratio",
    y     = "Prop. Stocking Paxlovid\n(Telehealth Sites)",
    title = "Simulation: OR, Sample Size, and Noise vs. Paxlovid Availability"
  ) +
  theme_bw(base_size = 9) +
  theme(legend.position  = "bottom",
        legend.key.size  = unit(0.3, "cm"),
        legend.text      = element_text(size = 6.5),
        legend.title     = element_text(size = 7, face = "bold"),
        legend.spacing.x = unit(0.15, "cm"),
        strip.background = element_rect(fill = "#deebf7"),
        strip.text       = element_text(size = 8),
        axis.text.x      = element_text(size = 6.5, angle = 45, hjust = 1),
        axis.text.y      = element_text(size = 7),
        axis.title       = element_text(size = 8),
        plot.title       = element_text(face = "bold", size = 8.5, hjust = 0.5),
        panel.spacing    = unit(0.4, "cm"))
```

---

# Interpretation

The simulations revealed several important patterns:

**Effect of the odds ratio:** As the OR increases from 1 (no association) to 80,
the simulated proportion of telehealth sites stocking Paxlovid rises rapidly and
approaches a plateau near 100% for high OR values. This plateau reflects that
probabilities are bounded between 0 and 1, a key constraint in binary outcome
modeling. The steep rise occurs between OR = 3 and OR = 20, confirming that the
observed OR (~30 in the real data) represents a very strong association.

**Effect of sample size:** At small sample sizes (n $\leq$ 500), estimates are
highly variable --- particularly for telehealth sites which are rare (~0.6% of the
sample per the Checkpoint 3 data), meaning many simulations yield zero telehealth
observations. As sample size grows toward the observed 64,237, estimates stabilize
and converge on the true parameter values. This underscores the importance of
adequate sample size when studying rare exposures.

**Effect of noise:** Increasing the noise parameter (log-odds SD = 0.5, 1.5)
introduces variability in the baseline probability $p_0$, which propagates to $p_1$
and widens the spread of simulated proportions. At high noise (SD = 1.5), even
large sample sizes show non-trivial variability, reflecting real-world
heterogeneity among provider sites. For power analyses, this suggests that
accounting for site-level noise is critical to avoid overestimating statistical
power.

**Implications for power analysis:** Future simulations (Final Report) should
prioritize sample sizes in the range of 1,000--10,000, as power will likely
plateau at larger n. Given the very large observed effect size, even moderate
sample sizes may be sufficient to detect the association --- but rare exposure
prevalence requires careful consideration of the expected number of telehealth
sites in any sub-sample.

---

# AI Use Disclosure Statement

Claude was used to assist in generating the R code for visualization. 
